Using Local PCA to Investigate Population Structure Along the Genome

### 2016-02-10 local_pca_proposal.tex
switched bibliography to natbib

\documentclass[11pt, oneside]{article}   	% use "amsart" instead of "article" for AMSLaTeX format
\usepackage{geometry}                		% See geometry.pdf to learn the layout options. There are lots.
\geometry{letterpaper}                   		% ... or a4paper or a5paper or ... 
%\geometry{landscape}                		% Activate for for rotated page geometry
%\usepackage[parfill]{parskip}    		% Activate to begin paragraphs with an empty line rather than an indent
\usepackage{graphicx}				% Use pdf, png, jpg, or eps§ with pdflatex; use eps in DVI mode
								% TeX will automatically convert eps --> pdf in pdflatex		
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{setspace}
\title{Using Local PCA to Investigate Population Structure Along the Genome}
\author{Han Li, Peter Ralph}
%\date{}							% Activate to display a given date or no date

@@

About 37 years ago, Cavalli-Sforza et al. first applied Principal component analysis (PCA) in population genetics to construct maps summarizing genetic variation \cite{key1}. Nowadays, PCA is a widely used powerful non-parametric method to extract information from genetic data. PCA results are derived from the covariance matrix of genotype matrix, which describes population structure, demographic history, and pedigree tree and so on. The results of PCA can be directly related to the underlying genealogical history of the samples, such as coalescence time (time to most recent common ancestor) and migration rate between populations \cite{key1,key2,key3}. Through dimension-reduction, PCA can identify key components of population structure, which describes how different samples are related, and plots of first two Principal components (PCs) can mimic the samples' geographic origin to some extent. The closeness of two samples in PC plots represents the similarity of their genetic information. Since population structure describes how different samples are related, and generally, samples living closer tends to be more genetically similar and thus tends to be clustered in PC plots \cite{key4,key6}.

@@

\subsection{Cut each genome into windows}
We cut each recoded matrix into sub matrixes by that have the same columns but fewer rows than the original matrix. Then apply Principal Component Analysis (PCA) on each window. Here's a brief summary about how is PCA carried out on genomic data \cite{key3}. Starting with the recoded genotype matrix Z, where Z is a L*N matrix ( L is SNP number ; N is sample size), then zero-center the matrix Z to X to make the data rows have equal variance. Get the covariance matrix of X (denote as matrix C) and compute the eigenvectors and eigenvalues of the covariance matrix C. The \textit{i}th principal component is the \textit{i}th eigenvector of C.

@@

\begin{multline}
Distance_{ij} \\ =\sqrt{\left \{\left ( V_{1}\cdot V_{1}^{} \right )^{2}+\left ( V_{2}\cdot V_{2}^{} \right )^{2}+\left ( U_{1}\cdot U_{1}^{} \right )^{2} +\left ( U_{2}\cdot U_{2}^{} \right )^{2} -2\left [ \left ( V_{1}\cdot U_{1}^{} \right )^{2}+\left ( V_{1}\cdot U_{2}^{} \right )^{2}+\left ( V_{2}\cdot U_{1}^{} \right )^{2} +\left ( V_{2}\cdot U_{2}^{} \right )^{2}\right ]\right \}}
\end{multline}

@@

Fig1.

@@

\noindent 2. Uneven sampling has a strong influence on PCA projections \cite{key3}. Our human data, POPRES, is unevenly sampled including 346 African-Americas, 73 Asians, 359 Indian Asians and 3187 Europeans. First, we'll try sub-sampling Europeans to balance the population size for the 4 population and repeat the process on the resampled data. Second, we'll try to apply the whole process on only European samples to see the genetic variation inside European samples. Third, we want to try different scheme of adding a weighting matrix to the covariance matrix of genotype data, thus to the reduce the influence of uneven sampling.

@@

\begin{thebibliography}{99}
\bibitem{key1} Menozzi, P., Piazza, A. \& Cavalli-Sforza, L. \lq\lq Synthetic maps of human gene frequencies in Europeans.'' Science 201, 786--792 (1978).
\bibitem{key2} Novembre, John, and Matthew Stephens. \lq\lq Interpreting principal component analyses of spatial population genetic variation." Nature genetics 40.5 (2008): 646-649.
\bibitem{key3} McVean, Gil. \lq\lq A genealogical interpretation of principal components analysis." PLoS Genet 5.10 (2009): e1000686.
\bibitem{key4} Novembre, John, et al. \lq\lq Genes mirror geography within Europe." Nature 456.7218 (2008): 98-101.
\bibitem{key5} Borg, Ingwer, and Patrick JF Groenen. Modern multidimensional scaling: Theory and applications. Springer Science \& Business Media, 2005.
\bibitem{key6} Patterson, N., A. L. Price, and D. Reich. \lq\lq Population structure and eigenanalysis." PLoS Genet 2.12 (2006): e190.
\bibitem{key7} Price, Alkes L., et al. \lq\lq Principal components analysis corrects for stratification in genome- wide association studies." Nature genetics 38.8 (2006): 904-909.
\bibitem{key8} Nelson, Matthew R., et al. \lq\lq The Population Reference Sample, POPRES: a resource for population, disease, and pharmacological genetics research." The American Journal of Human Genetics 83.3 (2008): 347-358.
\bibitem{key9} Efron, Bradley, and B. Efron. The jackknife, the bootstrap and other resampling plans. Vol. 38. Philadelphia: Society for industrial and applied mathematics, 1982.
\bibitem{key10} Engelhardt, Barbara E., and Matthew Stephens. \lq\lq Analysis of population structure: a unifying framework and novel methods based on sparse factor analysis." PLoS Genet 6.9 (2010): e1001117.
\bibitem{key11} Corbett-Detig, Russell B., and Daniel L. Hartl. \lq\lq Population genomics of inversion polymorphisms in Drosophila melanogaster." (2012): e1003056.
\bibitem{key12} Antonacci, Francesca, et al. \lq\lq Characterization of six human disease-associated inversion polymorphisms." Human molecular genetics 18.14 (2009): 2555-2566.
\bibitem{key13} Ma, Jianzhong, and Christopher I. Amos. \lq\lq Investigation of inversion polymorphisms in the human genome using principal components analysis." (2012): e40224.
\bibitem{key14} Kulikova, Olga, et al. \lq\lq Satellite repeats in the functional centromere and pericentromeric heterochromatin of Medicago truncatula." Chromosoma 113.6 (2004): 276-283.
\bibitem{key15} Kulikova, Olga, et al. \lq\lq Integration of the FISH pachytene and genetic maps of Medicago truncatula." The Plant Journal 27.1 (2001): 49-58.
\bibitem{key16} Paape, Timothy, et al. \lq\lq Selection, genome-wide fitness effects and evolutionary rates in the model legume Medicago truncatula." Molecular ecology 22.13 (2013): 3525-3538.

\end{thebibliography}

>>>>

\documentclass[11pt, oneside]{article}   	% use "amsart" instead of "article" for AMSLaTeX format
\usepackage{geometry}                		% See geometry.pdf to learn the layout options. There are lots.
\geometry{letterpaper}                   		% ... or a4paper or a5paper or ... 
%\geometry{landscape}                		% Activate for for rotated page geometry
%\usepackage[parfill]{parskip}    		% Activate to begin paragraphs with an empty line rather than an indent
\usepackage{graphicx}				% Use pdf, png, jpg, or eps§ with pdflatex; use eps in DVI mode
								% TeX will automatically convert eps --> pdf in pdflatex		
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{setspace}
\usepackage{natbib} \bibpunct{(}{)}{;}{author-year}{}{,} 
\usepackage[hidelinks]{hyperref}

\title{Using Local PCA to Investigate Population Structure Along the Genome}
\author{Han Li, Peter Ralph}
%\date{}							% Activate to display a given date or no date

\usepackage{color}
\newcommand{\plr}[1]{{\em \color{blue} #1}}

@@

About 37 years ago, \citet{menozzi1978synthetic} first applied Principal component analysis (PCA) in population genetics to construct maps summarizing genetic variation \citep{menozzi1978synthetic}. Nowadays, PCA is a widely used powerful non-parametric method to extract information from genetic data. PCA results are derived from the covariance matrix of genotype matrix, which describes population structure, demographic history, and pedigree tree and so on. The results of PCA can be directly related to the underlying genealogical history of the samples, such as coalescence time (time to most recent common ancestor) and migration rate between populations \citep{menozzi1978synthetic,novembre2008interpreting,mcvean2009genealogical}. Through dimension-reduction, PCA can identify key components of population structure, which describes how different samples are related, and plots of first two Principal components (PCs) can mimic the samples' geographic origin to some extent. The closeness of two samples in PC plots represents the similarity of their genetic information. Since population structure describes how different samples are related, and generally, samples living closer tends to be more genetically similar and thus tends to be clustered in PC plots \citep{novembre2008europe,key6}.

@@

\subsection{Cut each genome into windows}
We cut each recoded matrix into sub matrixes by that have the same columns but fewer rows than the original matrix. Then apply Principal Component Analysis (PCA) on each window. Here's a brief summary about how is PCA carried out on genomic data \cite{mcvean2009genealogical}. Starting with the recoded genotype matrix Z, where Z is a L*N matrix ( L is SNP number ; N is sample size), then zero-center the matrix Z to X to make the data rows have equal variance. Get the covariance matrix of X (denote as matrix C) and compute the eigenvectors and eigenvalues of the covariance matrix C. The \textit{i}th principal component is the \textit{i}th eigenvector of C.

@@

\plr{changed "Distance" to $D$}
\begin{align}
    \begin{split}
    D_{ij} &= 
        \left \{
                \left ( 
                    V_{1}\cdot V_{1}^{} 
                \right )^{2}
                + \left ( V_{2}\cdot V_{2}^{} \right )^{2}
                + \left ( U_{1}\cdot U_{1}^{} \right )^{2} 
                +\left ( U_{2}\cdot U_{2}^{} \right )^{2} 
        \right. \\ 
        & \left. \qquad {}  
            -2\left [ 
                \left ( V_{1}\cdot U_{1}^{} \right )^{2}
                + \left ( V_{1}\cdot U_{2}^{} \right )^{2}
                + \left ( V_{2}\cdot U_{1}^{} \right )^{2} 
                + \left ( V_{2}\cdot U_{2}^{} \right )^{2}
            \right ]
        \right \}^{1/2}
    \end{split}
\end{align}

@@

\begin{figure}
    \begin{center}
       % \includegraphics[width=0.7\textwidth]{figs/example_cline}
    \end{center}
    \caption{
        The caption goes here.
        \label{fig:pde_clines}
    }
\end{figure}

@@

\noindent 2. Uneven sampling has a strong influence on PCA projections \cite{mcvean2009genealogical}. Our human data, POPRES, is unevenly sampled including 346 African-Americas, 73 Asians, 359 Indian Asians and 3187 Europeans. First, we'll try sub-sampling Europeans to balance the population size for the 4 population and repeat the process on the resampled data. Second, we'll try to apply the whole process on only European samples to see the genetic variation inside European samples. Third, we want to try different scheme of adding a weighting matrix to the covariance matrix of genotype data, thus to the reduce the influence of uneven sampling.

@@

\bibliographystyle{plainnat}
\bibliography{references}  % if bibtex file is called references.bib

% \begin{thebibliography}{99}
% \bibitem{menozzi1978synthetic} Menozzi, P., Piazza, A. \& Cavalli-Sforza, L. \lq\lq Synthetic maps of human gene frequencies in Europeans.'' Science 201, 786--792 (1978).
% \bibitem{novembre2008interpreting} Novembre, John, and Matthew Stephens. \lq\lq Interpreting principal component analyses of spatial population genetic variation." Nature genetics 40.5 (2008): 646-649.
% \bibitem{mcvean2009genealogical} McVean, Gil. \lq\lq A genealogical interpretation of principal components analysis." PLoS Genet 5.10 (2009): e1000686.
% \bibitem{novembre2008europe} Novembre, John, et al. \lq\lq Genes mirror geography within Europe." Nature 456.7218 (2008): 98-101.
% \bibitem{key5} Borg, Ingwer, and Patrick JF Groenen. Modern multidimensional scaling: Theory and applications. Springer Science \& Business Media, 2005.
% \bibitem{key6} Patterson, N., A. L. Price, and D. Reich. \lq\lq Population structure and eigenanalysis." PLoS Genet 2.12 (2006): e190.
% \bibitem{key7} Price, Alkes L., et al. \lq\lq Principal components analysis corrects for stratification in genome- wide association studies." Nature genetics 38.8 (2006): 904-909.
% \bibitem{key8} Nelson, Matthew R., et al. \lq\lq The Population Reference Sample, POPRES: a resource for population, disease, and pharmacological genetics research." The American Journal of Human Genetics 83.3 (2008): 347-358.
% \bibitem{key9} Efron, Bradley, and B. Efron. The jackknife, the bootstrap and other resampling plans. Vol. 38. Philadelphia: Society for industrial and applied mathematics, 1982.
% \bibitem{key10} Engelhardt, Barbara E., and Matthew Stephens. \lq\lq Analysis of population structure: a unifying framework and novel methods based on sparse factor analysis." PLoS Genet 6.9 (2010): e1001117.
% \bibitem{key11} Corbett-Detig, Russell B., and Daniel L. Hartl. \lq\lq Population genomics of inversion polymorphisms in Drosophila melanogaster." (2012): e1003056.
% \bibitem{key12} Antonacci, Francesca, et al. \lq\lq Characterization of six human disease-associated inversion polymorphisms." Human molecular genetics 18.14 (2009): 2555-2566.
% \bibitem{key13} Ma, Jianzhong, and Christopher I. Amos. \lq\lq Investigation of inversion polymorphisms in the human genome using principal components analysis." (2012): e40224.
% \bibitem{key14} Kulikova, Olga, et al. \lq\lq Satellite repeats in the functional centromere and pericentromeric heterochromatin of Medicago truncatula." Chromosoma 113.6 (2004): 276-283.
% \bibitem{key15} Kulikova, Olga, et al. \lq\lq Integration of the FISH pachytene and genetic maps of Medicago truncatula." The Plant Journal 27.1 (2001): 49-58.
% \bibitem{key16} Paape, Timothy, et al. \lq\lq Selection, genome-wide fitness effects and evolutionary rates in the model legume Medicago truncatula." Molecular ecology 22.13 (2013): 3525-3538.
% \end{thebibliography}
